An Empirical Analysis of the Role of Energy in Economic Growth

### 2015-01-03 Articles/Article_EnerEcon_2014/EconGrowth3.Rnw
Edits to discussion of useful work.

The inclusion of energy as a factor of production 
in Cobb-Douglas and CES production functions 
is generally accepted by most 
resource and energy economists
\citep{Smulders:2003wc, vanderZwaan:2002wn, Kemfert:1998vd}.
Researchers interested in explaining long-term economic growth 
and possible constraints on growth 
have started to include energy as an explicit factor of production
\citep{Froling:2009vo, Tahvonen:2001tz}.  
An alternative approach, attempting to be 
``\dots more consistent with physical laws \dots'' \citep{Ayres:2001tx}
and classified by \citet{Stern:2011bq} as an ecological economic model of growth, 
is the LINEX model, 
wherein energy is the only factor of production,
and capital~($k$), labor~($l$), and energy~($e$)
combine to provide efficiencies that augment energy.
Technological change is manifest as increased efficiency:
more useful work done per joule of exergy \citep{Ayres:2005kn}.
There is no Solow-style term ($\mathrm{e}^{\lambda t}$) for technological change.

Unsatisfactorily, both the choice of a production function 
and the parameters used to populate these functions 
affect results significantly.
This situation leaves far too much latitude 
for misuse of model specifications 
and selection of model parameters 
to support certain theories (rather than empirically testing the theories
and parameters \emph{before} they are employed).  
For example, \citet{Lecca:2011kr} pointed out that 
the modellers' choice for the structure of a production function 
reflects his or her assumptions about
the elasticities among inputs, and 
``\dots the point at which energy enters the [nested CES] production function 
has a significant impact on CGE [computable general equilibrium] model results.''
\citet{Duffy:2000ty} rejected the Cobb-Douglas production function altogether, 
finding that capital and labour are more substitutable 
in richer countries and less substitutable in poorer countries 
than Cobb-Douglas implies.  
However, \citet{Miller:2008vd} points to the 
``\dots reasonably accurate long-term economic forecasts,'' 
achieved with the use of Cobb-Douglas production functions in economic models.  
\citet{Shen:2013vp} provide a recent example 
wherein empirical analysis on the estimated substitution elasticities 
is performed before choosing any particular 
nested CES production function structure. 
As a result, \citeauthor{Shen:2013vp} propose an atypical CES nesting structure 
for the Chinese economy. **** More detail here about nesting? --MKH ****  
\citet{Saunders:2008aa} evaluated several production and cost functions
(including the Cobb-Douglas, CES, and LINEX functions studied herein, 
see Section~\ref{sec:Prod_Func_Forms})
to assess whether they are capable of reproducing 
rebound and backfire effects.%
  \footnote{
  See \citet{Sorrell:2007um} for details about rebound and backfire effects.
  }
\citeauthor{Saunders:2008aa} concludes that,
of the production functions evaluated in the present study,
only the CES function is a candidate for 
rebound analysis.
The Cobb-Douglas function was found to exhibit only backfire.
The LINEX function was found to 
``not satisfy the standard concavity (diminishing marginal productivity) 
conditions.''~\citet[p.~2195]{Saunders:2008aa}

In addition to model structure, 
the way in which model parameters are determined is also subject to debate.
The \emph{cost share theorem} from mainstream economics states that 
output elasticities 
$\alpha$, $\beta$, and $\gamma$ 
(if augmenting the production function by energy)
can be calibrated from national accounts data as the 
cost shares of capital, labor, and energy, respectively.
However, the cost share theorem applies only under conditions of perfect competition
in the markets for the factors of production.
Although no market ever exhibits perfect competition, 
in our opinion, energy markets are \emph{least} likely to be ``perfect.''
Many governments have parastatal or full control 
of electricity generation and/or fuel markets.  
Other governments subsidize or incentivize energy markets.  
And, there are always health and environmental externalities 
that are not priced into fossil fuel energy markets.  
If energy markets are farther from ``perfect'' 
than capital and/or labor markets, 
then energy is the factor of production least likely to have its
output elasticity~($\gamma$) be well-represented 
by cost share as reported in national accounts.
Thus, when augmenting production functions by energy, 
much care should be taken in estimating 
the relative importance of the factors of production 
($\alpha$, $\beta$, and $\gamma$). 
\citet{Kummel:2010vz} showed that the cost share theorem applies 
only when economic growth is unconstrained.
Real-world constraints such as the degree of automation and
the degree of utilization of capital stock 
result in shadow prices that move the values 
of $\alpha$, $\beta$, and $\gamma$ away from 
their cost shares in the economy.~\citep[Equations~12 and~13]{Kummel:2010vz}
Thus, for this study, 
we reject the cost share theorem and
we have no \emph{a-priori} expectation that output elasticities 
estimated from historical output data
will match cost shares.

Thus, both the choice of production function structure 
and the values of the model parameters 
drive model results, and empirical and statistical testing 
of such approaches is prudent, but rarely conducted.

These are not idle academic debates. 
Macroeconomic growth models are used often to inform economic policy,
and the paths of economies worldwide are determined by
such policy choices. 
Therefore, we agree with those who express serious concerns 
about the reliability of the results from economic growth models.~\citep{Lecca:2011kr} 
We understand, however, 
that macroeconomic policymaking is needed 
and is increasingly informed by integrated energy-economy models, 
in a world where energy constraints are possible.~\citep{Beckman:2011gb, 
Benes:2012tm, Nel:2010fv, Kummel:2010vz}
Unfortunately, there has been little empirical work assessing the predictive ability 
of various energy-augmented production functions for different types of economies. 
And, very few studies of the importance of energy in economic growth include 
rigorous statistical assessments of energy-related fitting parameters. 
Thus, we support the call from \citet{Shen:2013vp} that both
the structure of the production function 
and the parameters that populate it 
need empirical testing. 
Building on the argument of \citet{Stern:2011bq} 
that energy may or may not be required in production functions 
depending on its abundance or scarcity, 
we will test whether different types of countries 
may need different factors of production to explain their respective growth paths, 
leaving open the possibility for 
different explanations of growth in various countries
at different stages of their development, 
a central point of long-wave economic growth modeling.~\citep{Froling:2009vo, Jones:2001wn, Galor:2000up}

This paper contributes to the literature on the role of energy in economic growth
by applying statistically rigorous 
methods along two coordinates of analysis: 
form of the energy augmented production function 
and type of economy.

The remainder of this paper is organized as follows. 
% Section~\ref{sec:lit_review} provides a necessarily brief overview of relevant literature.
Section~\ref{sec:Coordinates} discusses the coordinates of analysis for the present study. 
Section~\ref{sec:Sources_of_Data} details sources of data and calculations to 
generate time series for economic output ($y$) and factors of production ($k$, $l$, and $q$). 
The procedures for and results from parameter estimation for the various production functions 
are discussed in Section~\ref{sec:Parameter_Estimation}. 
Section~\ref{sec:Discussion} discusses implications of the parameter estimation 
results, and Section~\ref{sec:Conclusion} concludes.

@@

The LINEX model is intriguing because of its thermodynamic roots, 
but our implementation suffers from 
poor reproduction of historical data and
difficulty in respecting parameter constraints.
If the LINEX function is to be used, we recommend that 
\citeapos{Kummel:2011aa} fitting approach (logistic curves for $a_0$ and $c_t$) be employed
and that the results be thoroughly tested and compared against other production functions.

>>>>

The inclusion of energy as a factor of production 
in Cobb-Douglas and CES production functions 
is generally accepted by most 
resource and energy economists \citep{Smulders:2003wc, vanderZwaan:2002wn, Kemfert:1998vd}.
Researchers interested in explaining long-term economic growth 
and possible constraints on growth 
have begun to include energy 
as an explicit factor of production, too \citep{Froling:2009vo, Tahvonen:2001tz}.

An alternative approach, attempting to be 
``\dots more consistent with physical laws \dots'' \citep{Ayres:2001tx}
and classified by \citet{Stern:2011bq} as an ecological economic model of growth, 
is the LINEX model, 
wherein energy is the only factor of production,
and capital~($k$), labor~($l$), and energy~($e$)
combine to provide efficiencies that augment energy.
There is no Solow-style term ($\mathrm{e}^{\lambda t}$) for technological change.
Rather, technological change is manifest as increased efficiency:
more useful work done per joule of exergy input to the economy \citep{Ayres:2005kn}.
\citet{Kummel:1980wx} developed the LINEX function
using energy quantified as primary thermal energy input to the economy.
\citeauthor{Kummel:1980wx} represented 
energy efficiency coefficients in the LINEX function by logistic functions,
effectively converting from primary thermal energy to energy at the point of use
within the LINEX function itself.

In a series of papers culminating in a book~\citep{Ayres:2010ug},
\citeauthor{Ayres:2010ug} used the LINEX function with energy quantified as ``useful work,''
which is much closer to the point of use than 
the primary thermal energy quantification favored by \citeauthor{Kummel:1980wx}.
\citeauthor{Ayres:2010ug} do not find the need to use logistic functions representing energy efficiency,
presumably because useful work ``already incorporate[s] the energy-related efficiency changes 
in the economic system''~\citep[p.~35]{Kummel:2013aa}.
Both \citeauthor{Kummel:1980wx} and \citeauthor{Ayres:2010ug}
report satisfactory agreement between the LINEX function and historical economic output.

Unsatisfactorily, both the choice of a production function 
and the parameters used to populate these functions 
affect results significantly.
This situation leaves far too much latitude 
for misuse of model specifications 
and selection of model parameters 
to support certain theories, rather than empirically testing the theories
and parameters \emph{before} they are employed.  
For example, \citet{Lecca:2011kr} pointed out that 
the modellers' choice for the structure of a production function 
reflects his or her assumptions about
the elasticities among inputs, and 
``\dots the point at which energy enters the [nested CES] production function 
has a significant impact on CGE [computable general equilibrium] model results.''
\citet{Duffy:2000ty} rejected the Cobb-Douglas production function altogether, 
finding that capital and labour are more substitutable 
in richer countries and less substitutable in poorer countries 
than Cobb-Douglas implies.  
However, \citet{Miller:2008vd} points to the 
``\dots reasonably accurate long-term economic forecasts,'' 
achieved with the use of Cobb-Douglas production functions in economic models.  
\citet{Shen:2013vp} provide a recent example 
wherein empirical analysis on the estimated substitution elasticities 
is performed before choosing any particular 
nested CES production function structure. 
As a result, \citeauthor{Shen:2013vp} propose an atypical CES nesting structure 
for the Chinese economy. **** Martin: More detail here about nesting? --MKH ****  
\citet{Saunders:2008aa} evaluated several production and cost functions
(including the Cobb-Douglas, CES, and LINEX functions studied herein, 
see Section~\ref{sec:Prod_Func_Forms})
to assess whether they are capable of reproducing 
rebound and backfire effects.%
  \footnote{
  See \citet{Sorrell:2007um} for background on rebound and backfire.
  }
\citeauthor{Saunders:2008aa} concludes that,
of the production functions evaluated in the present study,
only the CES function is a candidate for 
rebound analysis.
The Cobb-Douglas function was found to exhibit only backfire.
The LINEX function was found to 
``not satisfy the standard concavity (diminishing marginal productivity) 
conditions.''~\citet[p.~2195]{Saunders:2008aa}

In addition to model structure, 
the way in which model parameters are determined is also subject to debate.
The \emph{cost share theorem} from mainstream economics states that 
output elasticities 
can be calibrated from national accounts data as the 
cost shares of factors of production: capital, labor, 
and energy (if augmenting the production function by energy).
However, the cost share theorem applies only under conditions of perfect competition
in the markets for the factors of production.
Although no market ever exhibits perfect competition, 
in our opinion energy markets are \emph{least} likely to be ``perfect.''
Many governments have parastatal or full control 
of electricity generation and fuel markets.  
Other governments subsidize or incentivize energy markets.  
And, there are always health and environmental externalities 
that are not priced into fossil fuel energy markets.  
If energy markets are farther from ``perfect'' 
than capital and labor markets, 
then energy is the factor of production least likely to have its
output elasticity be well-represented 
by cost share as reported in national accounts~\citep{Ayres:2013aa, Kummel:2013aa}.
Thus, when augmenting production functions by energy, 
much care should be taken in estimating 
the relative importance of the factors of production. 
\citet{Kummel:2010vz} showed that the cost share theorem applies 
only when economic growth is unconstrained.
Real-world constraints such as the degree of automation and
the degree of utilization of capital stock 
result in shadow prices that move 
output elasticities away from 
their cost shares in the economy \citep[Equations~12 and~13]{Kummel:2010vz}.
Thus, for this study, 
we reject the cost share theorem, and
we have no \emph{a-priori} expectation that output elasticities 
estimated from historical output data
will match cost shares.

Thus, there are stong indications in the literature that
both the choice of production function structure 
and the values of the model parameters 
drive model results, and empirical and statistical testing 
of such approaches is prudent, but rarely conducted.

These are not idle academic debates. 
Macroeconomic growth models are used often to inform economic policy,
and the paths of economies worldwide are determined by
such policy choices. 
Therefore, we agree with those who express serious concerns 
about the reliability of the results from economic growth models \citep{Lecca:2011kr}.
We understand, however, 
that macroeconomic policymaking is needed 
and is increasingly informed by integrated energy-economy models, 
in a world where energy constraints are possible \citep{Beckman:2011gb, 
Benes:2012tm, Nel:2010fv, Kummel:2010vz}.
Unfortunately, there has been little empirical work assessing the predictive ability 
of various energy-augmented production functions for different types of economies. 
And, very few studies of the importance of energy in economic growth include 
rigorous statistical assessment of energy-related fitting parameters.

This paper contributes to the literature on the role of energy in economic growth
by applying statistically rigorous 
methods along two coordinates of analysis: 
form of the energy augmented production function 
and type of economy.
We support the call from \citet{Shen:2013vp} that both
the structure of the production function 
and the parameters that populate it 
need empirical testing. 
Building on the argument of \citet{Stern:2011bq} 
that energy may or may not be required in production functions 
depending on its abundance or scarcity, 
we will test whether different types of countries 
may need different factors of production to explain their respective growth paths, 
leaving open the possibility for 
different explanations of growth in various countries
at different stages of their development, 
a central point of long-wave economic growth modeling \citep{Froling:2009vo, Jones:2001wn, Galor:2000up}.

The remainder of this paper is organized as follows. 
Section~\ref{sec:Coordinates} discusses the coordinates of analysis for the present study. 
Section~\ref{sec:Sources_of_Data} details sources of data and calculations to 
generate time series for economic output ($y$) and factors of production ($k$, $l$, and $q$). 
The procedures for and results from parameter estimation for the various production functions 
are discussed in Section~\ref{sec:Parameter_Estimation}. 
Section~\ref{sec:Discussion} discusses implications of the parameter estimation 
results, and Section~\ref{sec:Conclusion} concludes.

@@

The LINEX model is intriguing because of its thermodynamic roots, 
but our implementation suffers from 
poor reproduction of historical data and
difficulty in respecting parameter constraints.
If the LINEX function is to be used, 
we recommend that \citeapos{Kummel:2011aa} fitting approach 
(energy quantified by primary thermal energy and logistic curves for $a_0$ and $c_t$) 
or \citeapos{Ayres:2010ug} approach (energy quantified by useful work and constant values for $a_0$ and $c_t$) 
be employed 
and that the results be 
thoroughly tested and compared 
against other production functions.
